Theory of surface second-harmonic generation for semiconductors including effects of nonlocal operators

### 2014-12-04 sent/manuscript.tex
sent to prb

A convenient way to accomplish the separation of the SH signal of
either surface is to introduce a ``cut function'', ${\mathbf{\cal C}}(z)$, which is 
usually taken to be unity over one half of the slab and zero over 
the other half.\cite{reiningPRB94}
In this case ${\mathbf{\cal C}}(z)$ will give the contribution of the 
side of the slab for which ${\mathbf{\cal C}}(z)=1$. 
As was done for the linear response,\cite{mendozaPRB06}
we can generalize this 
simple choice for ${\mathbf{\cal C}}(z)$ by a top-hat cut function
${\mathbf{\cal C}}^\ell(z)$ that selects a given layer,
\begin{equation}
{\mathbf{\cal C}}^\ell(z)=\Theta(z-z_\ell+\Delta_\ell^{b})  
            \Theta(z_\ell-z+\Delta_\ell^f),
\label{sz}
\end{equation} 
where $\Theta$ is the Heaviside function. Here, $\Delta_\ell^{f/b}$
is the distance that the $\ell$-th layer extends towards the front
$(f)$ or back $(b)$ from its $z_\ell$ position. 
We take $z_\ell$ to be at the center of an atom that 
belongs to layer $\ell$, so the previous
equation would give the $\ell$-th atomic-layer 
contribution to the nonlinear optical response.
$\Delta_\ell^f+\Delta_\ell^b$ is the thickness of layer $\ell$ 
(see Fig.~\ref{fslab}).
\begin{figure}
\centering
\includegraphics[scale=.7]{images/slab}
\caption{(color on line) A sketch of the super-cell. 
The atomic slab corresponds to the
circles representing the atoms of the system.
\label{fslab}} 
\end{figure}

@@

In this section we present a 
relevant test case to check the
consistency of our approach. We have selected
a clean Si(001) surface with a $2\times 1$ surface reconstruction.
The slab for such a surface could be chosen to be centrosymmetric 
by creating the front and back
surfaces with the same $2\times 1$
reconstruction. However, we choose to terminate one 
of the surfaces with hydrogen producing an ideal  
terminated bulk Si surface. The H atoms simply saturate the 
dangling bonds of the bulk-like Si atoms at the surface, as seen in 
Fig.~\ref{si2x1}.
We take the $z$ coordinate pointing out of the surface and the $x$
coordinate along the crystallographic [011] direction is parallel to the dimmers. 
\begin{figure}
\centering 
\includegraphics[scale=.8]{images/si2x1-crop}
\caption{(color on line) The slab shows a 
    clean Si(001)$2\times 1$ front surface 
    with an ideal terminated Si bulk back surface. The
    dangling bonds are H (small balls) saturated. 
    This image depicts 12 Si atomic layers
    with one H atomic layer. 
\label{si2x1}} 
\end{figure} 
The idea behind this slab configuration is that the 
cristalline symmetry of the H terminated surface imposes that 
$\chi_{\mathrm{H}}^{xxx}=0$.
 The $2\times 1$ surface has no 
such restrictions, so $\chi_{2\times 1}^{xxx}\ne 0$.
This is due to the fact that along the $y$ direction there is a mirror plane for the
H-saturated surface, whereas 
for the $2\times 1$ surface this mirror is lost as the dimers are
asymmetric along $x$. 
Thus, calculating $\chi^{xxx}$ for the full-slab, or the 
half-slab containing the $2\times 1$ surface\cite{note1}
should yield the same result since the contribution from the H
saturated surface is zero regardless. 
We must check that the following 
relationship is satisfied for this particular slab
\begin{equation*}
\chi_{\mathrm{half-slab}}^{xxx}(-2\omega;\omega,\omega) 
=
\chi_{\mathrm{full-slab}}^{xxx}(-2\omega;\omega,\omega) 
,
\end{equation*}
where
$\chi_{\mathrm{half-slab}}^{xxx}(-2\omega;\omega,\omega)$ is calculated using
${\mathbf{\cal C}}(z)=1$ from the upper half containing the $2\times 1$ 
surface reconstruction, as seen in Fig.~\ref{si2x1},
and $\chi_{\mathrm{full-slab}}^{xxx}(-2\omega;\omega,\omega)$ is calculated using
${\mathbf{\cal C}}(z)=1$ through the full slab.
We show the results for this comparison in the remainder  
of this section.  
Also, we checked that 
for the
dihydride surface
$\chi_{\mathrm{half-slab}}^{xxx}(-2\omega;\omega,\omega)=0$.

@@

Spin-orbit, local field, and electron-hole 
attraction\cite{beyond}
effects on the SHG process are all neglected.
Although these are important factors in the optical response of a semiconductor,
their efficient calculation is still theoretically and  
numerically challenging and  
under debate. This merits further study but is beyond the scope of this paper.
For a given slab size, we find the converged spectra 
to obtain the relevant parameters. The most important of 
these are: an energy cut-off of 10 Ha, an equal number of conduction and 
valence bands, and a set of 244 $\mathbf{k}$-points.
The $\mathbf{k}$-points are used for the linear analytic 
tetrahedron method for evaluating the
Brillouin zone integrals 
where special care was taken to examine the double resonances
of Eq.~\eqref{chis}. \cite{nastosPRB05}
All spectra were calculated with a Gaussian smearing of 0.15 eV.
\begin{figure}
\centering 
\includegraphics[scale=.8]{plots/fig1}
\caption{(color on line) 
$|\chi_{\mathrm{half-slab}}^{xxx}|$ vs $\hbar\omega$ 
for the slab
with 16, 24, 32, and 40 atomic Si layers. The front surface 
is in a clean $2\times 1$ reconstruction and the back 
surface is an ideal terminated bulk H-saturated dangling bonds (see Fig.~\ref{si2x1}).
\label{fig1}} 
\end{figure}
\begin{figure}
\centering 
\includegraphics[scale=.8]{plots/fig2}
\caption{(color on line) 
$\chi^{xxx}_{\mathrm{half-slab}}$
and 
$\chi^{xxx}_{\mathrm{full-slab}}$
vs $\hbar\omega$ for a slab with 32 
atomic Si layers plus one H layer. 
\label{fig2}} 
\end{figure}

@@

We proceed to explain 
some of the features seen in $|\chi^{xxx}_{2\times 1}|$ that, as
explained above, are obtained 
by calculating $|\chi^{xxx}_{\mathrm{half-slab}}|$.
First, from Fig.~\ref{fig2} we note a series of resonances 
that derive from 1$\omega$ and 2$\omega$ terms in
Eq.~\eqref{chis}. 
Notice that the 
2$\omega$ resonances start below $E_g/2$ where $E_g$ is the band gap
(0.53 eV  for LDA and
1.03 eV if the scissor is used with
$\hbar\Delta=0.5$ eV).
These resonances
come from the electronic states of the 
$2\times 1$ surface, that lie inside the bulk band gap of Si and are the 
well known electronic surface states.\cite{rohlfingPRB95} 
In Fig.~\ref{fig3}
we see that the effect of $\mathbf{v}^\mathrm{nl}$
 reduces the value of   
$|\chi^{xxx}_{2\times 1}|$  
by 15-20\% 
showing the importance of this contribution for a correct calculation  
of SSHG, in agreement with the analysis for bulk semiconductors.\cite{luppiPRB08}
However, the inclusion of $\mathbf{v}^\mathrm{nl}$ does not
changes the 
spectral shape of $|\chi^{xxx}_{2\times 1}|$;
 this also can be confirmed from
the cases of zero scissors correction from Fig.~\ref{fig2}.
\begin{figure}
\centering 
\includegraphics[scale=.8]{plots/fig3}
\caption{(color on line) 
$\chi^{xxx}_{2\times 1}$
vs $\hbar\omega$ for a slab with 32 
atomic Si layers plus one H layer, 
with and without the contribution from $\mathbf{v}^\mathrm{nl}$.
\label{fig3}} 
\end{figure}

To see the effect of the scissors correction, we take two different
finite values for $\hbar\Delta$. The first one
with a value of $\hbar\Delta=0.5$ eV, used in the above results, 
is the ``average'' GW gap taken from 
Ref.~\onlinecite{rohlfingPRB95} 
that is in agreement with Ref.~\onlinecite{garciaCPC01}. The second one
with a value of $\hbar\Delta=0.63$ eV is the ``average'' 
gap taken from Ref.~\onlinecite{asahiPRB00}, 
where more $\mathbf{k}$-points in the Brillouin zone were 
used to calculate its GW value.
From Fig.~\ref{fig4}
we note that the scissors correction 
shifts the spectra from it LDA value to higher energies as expected.
 However, contrary 
to the case of linear optics\cite{cabellosPRB09} the shift introduced 
by the scissors correction is not 
rigid, as pointed out in Ref.~\onlinecite{nastosPRB05}. 
 This is because the second-harmonic optical response mixes 
$1\omega$ and $2\omega$ transitions (see Eq.~\eqref{chis}), and accounts 
for the non-rigid shift. 
When we compare 
$|\chi^{xxx}_{2\times 1}|$ for the two finite values of $\hbar\Delta$,
we see that the first two peaks are almost rigidly 
shifted with a small difference in height while
the rest of the peaks are modified substantially. 
This behavior comes from the fact that the first two
peaks are almost exclusively related to the 
2$\omega$ resonances of
Eq.~\eqref{chis}. The other peaks are a combination 
of 1$\omega$ and 2$\omega$ resonances 
and yield a more varied spectrum.
We mention that for large gap materials, the 1$\omega$ and 2$\omega$ would be splited
showing a small interference effect, but still the 2$\omega$ would
strongly depend on the surface states.
 This way we see that small changes in the value of the 
scissors shift can in general affect the SSH susceptibility 
spectrum quite dramatically.
In Ref.~\onlinecite{adolphPRB00}, the authors already remarked
that nonlinear optical response of bulk materials
is more influenced by the electronic structure
of the material than the linear case. 
In the case of semiconducting surfaces the
problem is even more intricate due to the presence of electronic
surface states.
The high sensitivity of SSHG to the energy position  of surface
states, as seen in Fig.~\ref{fig4}, makes SSHG a good benchmark
spectroscopical tool for testing the validity of the inclusion of many-body effects,
and in particular the quasi-particle correction to the electronic states. 
\begin{figure}
\centering 
\includegraphics[scale=.8]{plots/fig4}
\caption{(color on line) 
$\chi^{xxx}_{2\times 1}$
vs $\hbar\omega$ for a slab with 32 
atomic Si layers plus one H layer, 
for two different values of 
the scissors correction $\hbar\Delta$. 
\label{fig4}} 
\end{figure}

>>>>

A convenient way to accomplish the separation of the SH signal of
either surface is to introduce a ``cut function'', ${\mathbf{\cal C}}(z)$, which is 
usually taken to be unity over one half of the slab and zero over 
the other half.\cite{reiningPRB94}
In this case ${\mathbf{\cal C}}(z)$ will give the contribution of the 
side of the slab for which ${\mathbf{\cal C}}(z)=1$. 
As was done for the linear response,\cite{mendozaPRB06}
we can generalize this 
simple choice for ${\mathbf{\cal C}}(z)$ by a top-hat cut function
${\mathbf{\cal C}}^\ell(z)$ that selects a given layer,
\begin{equation}
{\mathbf{\cal C}}^\ell(z)=\Theta(z-z_\ell+\Delta_\ell^{b})  
            \Theta(z_\ell-z+\Delta_\ell^f),
\label{sz}
\end{equation} 
where $\Theta$ is the Heaviside function. Here, $\Delta_\ell^{f/b}$
is the distance that the $\ell$-th layer extends towards the front
$(f)$ or back $(b)$ from its $z_\ell$ position. 
We take $z_\ell$ to be at the center of an atom that 
belongs to layer $\ell$, so the previous
equation would give the $\ell$-th atomic-layer 
contribution to the nonlinear optical response.
$\Delta_\ell^f+\Delta_\ell^b$ is the thickness of layer $\ell$ 
(see Fig.~\ref{fslab}).
\begin{figure}
\centering
\includegraphics[scale=.7]{fig1}
\caption{(color on line) A sketch of the super-cell. 
The atomic slab corresponds to the
circles representing the atoms of the system.
\label{fslab}} 
\end{figure}

@@

In this section we present a 
relevant test case to check the
consistency of our approach. We have selected
a clean Si(001) surface with a $2\times 1$ surface reconstruction.
The slab for such a surface could be chosen to be centrosymmetric 
by creating the front and back
surfaces with the same $2\times 1$
reconstruction. However, we choose to terminate one 
of the surfaces with hydrogen producing an ideal  
terminated bulk Si surface. The H atoms simply saturate the 
dangling bonds of the bulk-like Si atoms at the surface, as seen in 
Fig.~\ref{si2x1}.
We take the $z$ coordinate pointing out of the surface and the $x$
coordinate along the crystallographic [011] direction is parallel to the dimmers. 
\begin{figure}
\centering 
\includegraphics[scale=.8]{fig2}
\caption{(color on line) The slab shows a 
    clean Si(001)$2\times 1$ front surface 
    with an ideal terminated Si bulk back surface. The
    dangling bonds are H (small balls) saturated. 
    This image depicts 12 Si atomic layers
    with one H atomic layer. 
\label{si2x1}} 
\end{figure} 
The idea behind this slab configuration is that the 
cristalline symmetry of the H terminated surface imposes that 
$\chi_{\mathrm{H}}^{xxx}=0$.
 The $2\times 1$ surface has no 
such restrictions, so $\chi_{2\times 1}^{xxx}\ne 0$.
This is due to the fact that along the $y$ direction there is a mirror plane for the
H-saturated surface, whereas 
for the $2\times 1$ surface this mirror is lost as the dimers are
asymmetric along $x$. 
Thus, calculating $\chi^{xxx}$ for the full-slab, or the 
half-slab containing the $2\times 1$ surface\cite{note1}
should yield the same result since the contribution from the H
saturated surface is zero regardless. 
We must check that the following 
relationship is satisfied for this particular slab
\begin{equation*}
\chi_{\mathrm{half-slab}}^{xxx}(-2\omega;\omega,\omega) 
=
\chi_{\mathrm{full-slab}}^{xxx}(-2\omega;\omega,\omega) 
,
\end{equation*}
where
$\chi_{\mathrm{half-slab}}^{xxx}(-2\omega;\omega,\omega)$ is calculated using
${\mathbf{\cal C}}(z)=1$ from the upper half containing the $2\times 1$ 
surface reconstruction, as seen in Fig.~\ref{si2x1},
and $\chi_{\mathrm{full-slab}}^{xxx}(-2\omega;\omega,\omega)$ is calculated using
${\mathbf{\cal C}}(z)=1$ through the full slab.
We show the results for this comparison in the remainder  
of this section.  
Also, we checked that 
for the
dihydride surface
$\chi_{\mathrm{half-slab}}^{xxx}(-2\omega;\omega,\omega)=0$.

@@

Spin-orbit, local field, and electron-hole 
attraction\cite{beyond}
effects on the SHG process are all neglected.
Although these are important factors in the optical response of a semiconductor,
their efficient calculation is still theoretically and  
numerically challenging and  
under debate. This merits further study but is beyond the scope of this paper.
For a given slab size, we find the converged spectra 
to obtain the relevant parameters. The most important of 
these are: an energy cut-off of 10 Ha for the 16, 24, and 32 layered
slabs
and 13 Ha for the 40 layer slab,
an equal number of conduction and 
valence bands, and a set of 244 $\mathbf{k}$-points.
The $\mathbf{k}$-points are used for the linear analytic 
tetrahedron method for evaluating the
Brillouin zone integrals 
where special care was taken to examine the double resonances
of Eq.~\eqref{chis}. \cite{nastosPRB05}
All spectra were calculated with a Gaussian smearing of 0.15 eV.
\begin{figure}
\centering 
\includegraphics[scale=.8]{fig3}
\caption{(color on line) 
$|\chi_{\mathrm{half-slab}}^{xxx}|$ vs $\hbar\omega$ 
for the slab
with 16, 24, 32, and 40 atomic Si layers. The front surface 
is in a clean $2\times 1$ reconstruction and the back 
surface is an ideal terminated bulk H-saturated dangling bonds (see Fig.~\ref{si2x1}).
\label{fig1}} 
\end{figure}
\begin{figure}
\centering 
\includegraphics[scale=.8]{fig4}
\caption{(color on line) 
$\chi^{xxx}_{\mathrm{half-slab}}$
and 
$\chi^{xxx}_{\mathrm{full-slab}}$
vs $\hbar\omega$ for a slab with 32 
atomic Si layers plus one H layer. 
\label{fig2}} 
\end{figure}

@@

We proceed to explain 
some of the features seen in $|\chi^{xxx}_{2\times 1}|$ that, as
explained above, are obtained 
by calculating $|\chi^{xxx}_{\mathrm{half-slab}}|$.
First, from Fig.~\ref{fig2} we note a series of resonances 
that derive from 1$\omega$ and 2$\omega$ terms in
Eq.~\eqref{chis}. 
Notice that the 
2$\omega$ resonances start below $E_g/2$ where $E_g$ is the band gap
(0.53 eV  for LDA and
1.03 eV if the scissor is used with
$\hbar\Delta=0.5$ eV).
These resonances
come from the electronic states of the 
$2\times 1$ surface, that lie inside the bulk band gap of Si and are the 
well known electronic surface states.\cite{rohlfingPRB95} 
In Fig.~\ref{fig3}
we see that the effect of $\mathbf{v}^\mathrm{nl}$
 reduces the value of   
$|\chi^{xxx}_{2\times 1}|$  
by 15-20\% 
showing the importance of this contribution for a correct calculation  
of SSHG, in agreement with the analysis for bulk semiconductors.\cite{luppiPRB08}
However, the inclusion of $\mathbf{v}^\mathrm{nl}$ does not
changes the 
spectral shape of $|\chi^{xxx}_{2\times 1}|$;
 this also can be confirmed from
the cases of zero scissors correction from Fig.~\ref{fig2}.
\begin{figure}
\centering 
\includegraphics[scale=.8]{fig5}
\caption{(color on line) 
$\chi^{xxx}_{2\times 1}$
vs $\hbar\omega$ for a slab with 32 
atomic Si layers plus one H layer, 
with and without the contribution from $\mathbf{v}^\mathrm{nl}$.
\label{fig3}} 
\end{figure}

To see the effect of the scissors correction, we take two different
finite values for $\hbar\Delta$. The first one
with a value of $\hbar\Delta=0.5$ eV, used in the above results, 
is the ``average'' GW gap taken from 
Ref.~\onlinecite{rohlfingPRB95} 
that is in agreement with Ref.~\onlinecite{garciaCPC01}. The second one
with a value of $\hbar\Delta=0.63$ eV is the ``average'' 
gap taken from Ref.~\onlinecite{asahiPRB00}, 
where more $\mathbf{k}$-points in the Brillouin zone were 
used to calculate its GW value.
From Fig.~\ref{fig4}
we note that the scissors correction 
shifts the spectra from it LDA value to higher energies as expected.
 However, contrary 
to the case of linear optics\cite{cabellosPRB09} the shift introduced 
by the scissors correction is not 
rigid, as pointed out in Ref.~\onlinecite{nastosPRB05}. 
 This is because the second-harmonic optical response mixes 
$1\omega$ and $2\omega$ transitions (see Eq.~\eqref{chis}), and accounts 
for the non-rigid shift. 
When we compare 
$|\chi^{xxx}_{2\times 1}|$ for the two finite values of $\hbar\Delta$,
we see that the first two peaks are almost rigidly 
shifted with a small difference in height while
the rest of the peaks are modified substantially. 
This behavior comes from the fact that the first two
peaks are almost exclusively related to the 
2$\omega$ resonances of
Eq.~\eqref{chis}. The other peaks are a combination 
of 1$\omega$ and 2$\omega$ resonances 
and yield a more varied spectrum.
We mention that for large gap materials, the 1$\omega$ and 2$\omega$ would be splited
showing a small interference effect, but still the 2$\omega$ would
strongly depend on the surface states.
 This way we see that small changes in the value of the 
scissors shift can in general affect the SSH susceptibility 
spectrum quite dramatically.
In Ref.~\onlinecite{adolphPRB00}, the authors already remarked
that nonlinear optical response of bulk materials
is more influenced by the electronic structure
of the material than the linear case. 
In the case of semiconducting surfaces the
problem is even more intricate due to the presence of electronic
surface states.
The high sensitivity of SSHG to the energy position  of surface
states, as seen in Fig.~\ref{fig4}, makes SSHG a good benchmark
spectroscopical tool for testing the validity of the inclusion of many-body effects,
and in particular the quasi-particle correction to the electronic states. 
\begin{figure}
\centering 
\includegraphics[scale=.8]{fig6}
\caption{(color on line) 
$\chi^{xxx}_{2\times 1}$
vs $\hbar\omega$ for a slab with 32 
atomic Si layers plus one H layer, 
for two different values of 
the scissors correction $\hbar\Delta$. 
\label{fig4}} 
\end{figure}

### 2014-12-04 sent/submission_letter.tex
sent to prb

In this article we formulate a theoretical approach of surface
second-harmonic generation from semiconductor surfaces based on the
length gauge and the one electron density operator.  We calculate the
nonlinear second-order surface susceptibility tensor within the
independent particle approximation including for the first time in one
unique formulation (i) the scissors correction, needed to have the
correct value of the energy band gap, (ii) the contribution of the
nonlocal part of the pseudopotentials, routinely used in ab initio
band structure calculations, and (iii) the derivation for the
inclusion of the cut function, used to extract the surface response.
We test our scheme with a clean Si(001)2x1 reconstructed surface with
excellent results.  Therefore, the inclusion of the three
aforementioned contributions is very important and makes our scheme
unprecedented and opens the possibility to study surface
second-harmonic generation with more versatility and providing more
accurate results.  Most of the cited works appear in the Physical
Review Letters and Physical Review B, for which we think that this
article would have a good visibility in Physical Review B, as well.

>>>>

In this article we formulate a theoretical approach of surface
second-harmonic generation from semiconductor surfaces based on the
length gauge and the one electron density operator.  We calculate the
nonlinear second-order surface susceptibility tensor within the
independent particle approximation. We include for the first time in one
unique formulation: (i) the scissors correction, needed to have the
correct value of the energy band gap, (ii) the contribution of the
nonlocal part of the pseudopotentials, routinely used in ab initio
band structure calculations, and (iii) the derivation for the
inclusion of the cut function, used to extract the surface response.
We test our scheme with a clean Si(001)2x1 reconstructed surface with
excellent results.  Therefore, the inclusion of the three
aforementioned contributions is very important and makes our scheme
unprecedented and opens the possibility to study surface
second-harmonic generation with more versatility while providing more
accurate results.  Most of the cited works appear in the Physical
Review Letters and Physical Review B, for which we think that this
article would have good visibility in Physical Review B as well.
